CF-SuGAR Power Calculations

### 2025-03-20 Christine Chan/CF-SuGAR/power_calculations.qmd
PROC POWER calculations for Deb

---
title: "CF-SuGAR Power Calculations"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    fig-width: 10
    fig-height: 10
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
editor: source
---

@@

@@

annual$bmi_cat <- apply(annual, 1, function(r) {
  age <- as.numeric(r["Age_YrEnd"])
  gender <- as.character(r["Gender"])
  bmi <- as.numeric(r["bmi"])
  bmi_perc <- as.numeric(r["bmi_perc"])
  if (age <= 20) {
    return(cut(bmi_perc, c(-Inf, 10, 50, 85, Inf),
      right = F,
      labels = c("Undernourished", "At Risk", "Target/Goal", "Overweight")
    ))
  } else if (age > 20) {
    if (gender == "M") {
      return(cut(bmi, c(-Inf, 18.5, 23, 25, Inf),
        right = F,
        labels = c("Undernourished", "At Risk", "Target/Goal", "Overweight")
      ))
    } else if (gender == "F") {
      return(cut(bmi, c(-Inf, 18.5, 22, 25, Inf),
        right = F,
        labels = c("Undernourished", "At Risk", "Target/Goal", "Overweight")
      ))
annual$bmi_cat <- relevel(annual$bmi_cat, ref = "Target/Goal")

@@

```

# Registry

```{r}
# Origin for survival analyses
og <- ymd("2003-01-01")
# Need to use a dataframe not a tibble
annual <- as.data.frame(annual)
annual$tstart <- as.numeric(annual$tstart) - as.numeric(og)
annual$tstop <- as.numeric(annual$tstop) - as.numeric(og)
# Formula for base model
xvars <- c(
  "age_group", "Gender", "Race", "Hispanicrace", "bmi", "MutClass",
  "A_IsOnEnzymes", "tube_feeding", "liver_disease", "A_FEV1_pct_predicted",
  "NumPulmExacerbation", "A_corticosteroids1"
)
y <- "Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2)"
f <- as.formula(paste0(y, "~", paste0(xvars, collapse = "+")))
cox_res <- coxph(f, data = annual, cluster = eDWID)
cox.zph(cox_res)
# Forest plot and coefficients table
cox_res %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

## Base model with variables available in Lorenz et al.

```{r}
annual = annual%>% filter(Age_YrEnd<=22)
annual$age_group = cut(annual$Age_YrEnd,c(7,11,22))
# Formula for base model
xvars <- c(
  "age_group", "Gender"
)
y <- "Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2)"
f <- as.formula(paste0(y, "~", paste0(xvars, collapse = "+")))
cox_res <- coxph(f, data = annual, cluster = eDWID)
cox.zph(cox_res)
# Forest plot and coefficients table
cox_res %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

```{r}
# read in separate data files

@@

# create multogtt with time <4 years since there is little data beyond that
multogtt_reduced <- multogtt[multogtt$time <= 4, ]

##############
# COX MODELS #
##############
```

@@

```{r}
# CPH model for baseline 1 hour glucose
cox1hr <- coxph(
  Surv(Follow_up_time_yrs, CFRD) ~ age + Sex ,
  data = baseline_mult
)
# Forest plot and coefficients table
cox1hr %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox1hr,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

@@

```{r}
# CPH model for baseline 1 hour glucose
baseline_mult$age_group = cut(baseline_mult$age,c(7,11,22))

cox1hr <- coxph(
  Surv(Follow_up_time_yrs, CFRD) ~ age*Sex,
  data = baseline_mult
)

# Forest plot and coefficients table
cox1hr %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox1hr,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

# Notes

- I removed race from the models because the Lorenz et al. paper data only includes 8 non-white participants, so the coefficients exploded.

To Do

- Plot HbA1c by age for those who do develop CFRD
- Take out HbA1c for base model
- Include generalized R2 for the models
- Split age into equal tertiles
  2 x 3 table with rows for diabetes status and age as columns, cells are numbers

>>>>

---
title: "CF-SuGAR Power Calculations"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    fig-width: 10
    fig-height: 10
    toc: true
    toc-depth: 5
    toc-float: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
editor: source
---

@@

#| include: false

@@

@@

# Origin for survival analyses
og <- ymd("2003-01-01")
# Need to use a dataframe not a tibble
annual <- as.data.frame(annual)
annual$tstart <- as.numeric(annual$tstart) - as.numeric(og)
annual$tstop <- as.numeric(annual$tstop) - as.numeric(og)
# Remove ages > 22 to match Lorenz et al.
annual <- annual %>% filter(Age_YrEnd < 23)
# Lorenz et al. data cleaning
# read in separate data files

@@

# create multogtt with time <4 years since there is little data beyond that
multogtt_reduced <- multogtt[multogtt$time <= 4, ]
# rename data
lorenz <- baseline_mult
registry <- annual
rm(annual, baseline_mult)
```

# Data cleaning

- Participants over age 23 were excluded from the registry data, along with data collected prior to 2010.
- Age groups were chosen based on the Lorenz et al. data. We split the age column into equal tertiles:

```{r}
lorenz$age_group <- cut(lorenz$age, c(7, 10.48, 13.3, 22.2), right = F)
table(lorenz$age_group, lorenz$CFRD)
registry$age_group <- cut(registry$Age_YrEnd, c(7, 10.48, 13.3, 22.2), right = F)
```

# Registry base model

```{r}
registry_base <- coxph(
  Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2) ~
    age_group + Gender,
  data = registry, cluster = eDWID
)
# Forest plot and coefficients table
registry_base %>%
  tbl_regression(exponentiate = T)
ggcoef_table(registry_base,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

The above model had a generalized $R^2$ of `r as.numeric(summary(registry_base)$rsq[1])`.

@@

```{r}
# CPH model for baseline 1 hour glucose
lorenz_base <- coxph(
  Surv(Follow_up_time_yrs, CFRD) ~ age_group + Sex,
  data = lorenz
)
# Forest plot and coefficients table
lorenz_base %>%
  tbl_regression(exponentiate = T)
ggcoef_table(lorenz_base,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

The above model had a generalized $R^2$ of `r as.numeric(summary(lorenz_base)$rsq[1])`.

@@

```{r}
# CPH model for baseline 1 hour glucose
lorenz_1hr <- coxph(
  Surv(Follow_up_time_yrs, CFRD) ~ age_group + Sex + Baseline_1hr_glucose,
  data = lorenz
)
summary(lorenz_1hr)
# Forest plot and coefficients table
lorenz_1hr %>%
  tbl_regression(exponentiate = T)
ggcoef_table(lorenz_1hr,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

The above model had a generalized $R^2$ of `r as.numeric(summary(lorenz_1hr)$rsq[1])`.

# PROC POWER

## HR for 1 hour OGTT glucose

```{r}
exp(lorenz_1hr$coefficients["Baseline_1hr_glucose"] * 0.5)
exp(lorenz_1hr$coefficients["Baseline_1hr_glucose"])
exp(lorenz_1hr$coefficients["Baseline_1hr_glucose"] * 2)
```

## Regression of age group and sex on 1 hour OGTT glucose

```{r}
lin_mod <- lm(Baseline_1hr_glucose ~ age_group + Sex, data = lorenz)
summary(lin_mod)$r.squared
```

## SD of 1 hour OGTT glucose

```{r}
sd(lorenz$Baseline_1hr_glucose)
```

## Event probability

```{r}
registry %>%
  group_by(eDWID) %>%
  summarise(cfrd = unique(first_cfrd), .groups = "drop") %>%
  summarise(cfrd = mean(!is.na(cfrd)))
mean(lorenz$CFRD == 1)
```

# To Do
